\chapter{Social Networks}

When Rishnak was talking with his other ghost friends about his mathematical adventures with Ajur. Just then Rishnak spotted Ajur and Jura walking. Rishnak cut short his conversation with his friends and joined Ajur and Jura. Rishnak said that graph theory has been used in a lot of applications over the last century like in Engineering, Physics and Optimization Problems. However, with the prolific use of social network applications such as Facebook, Twitter and Linked in, Sociologists and psychologists have found a lot of applications using Graph Theory.  
Stanley Milgram, a famous psychologist wanted to show that the people are well connected. He devised an experiment where in he selected a collection of people in midwestern United States and asked them to send post cards to a person in Boston. If they do not the person then they could send the post card to a person that knows some one who knows the end person in Boston. The result of the experiment showed that most of the letters reached the destination person in Boston in 5 or 6 steps. The experiments showed the diameter (longest of the shortest distance between vertices) of a social network is very small.

Social network has also been used in connection with actors in an actor network. Actors/Actresses form the vertices and vertices are adjacent if the corresponding actors/actresses have acted in the same film. There is an actor by the name of Kevin Bacon who has acted with a number of actors/actresses. Questions like what is your Kevin Bacon distance were answered using Social networks. It is essentially the length of the path of in the co-actor network. A dedicated website finds a path \url{http://oracleofbacon.org/movielinks.php} There is even a play about six degrees of separation. Again this shows the diameter of the network is very small and there is a vertex with a large degree (Kevin Bacon).

Social networks have also been used to co-author (of scientific and mathematical publications) network. One of the famous 20-th Century Mathematician and a prolific contributor to Graph Theory Paul Erd\H{o}s has written many authors with a lot of co-authors. Authors become vertices and two vertices(authors) are connected by an edge if they have written a paper together. American Mathematical Society has a website that calculates Erd\H{o}s distance to authors
\url{https://mathscinet.ams.org/mathscinet/collaborationDistance.html}. Usually Erd\h{o}s distance to many graph theorists is a very small number because of the collaborative nature of the work and an author like Paul Erd\H{o}s has written a large number of papers with a large number of co-authors (vertex corresponding to Erd\H{o}s has a large degree).

Social networks provide a number of advantages:
\begin{enumerate}
    \item Promoting your ideas and products. You share your ideas to your friends. They in turn share this.
    So you may be able to influence others of your opinions. You will 
\end{enumerate}
